---
title: "NYPD Shooting"
author: "Michael Kilgore"
date: "2023-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Racial characteristics of NY Shooting Data

We start by pulling in the City of New York shooting data:

```{r importData}
library(tidyverse)
shooting_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
clean_shot = select(shooting_data, BORO, VIC_RACE)
clean_shot <- clean_shot %>% rename(bor = BORO, v_race = VIC_RACE)
clean_shot$bor = as.factor(clean_shot$bor)
clean_shot$v_race = as.factor(clean_shot$v_race)
```

### Data Summary and modifications
We see how racial information was recorded in a data summary, 
then modify it to match
the "Asian - White - Hispanic - Black" demo information from the
Furman Center data, which we'll import below.

```{r simplify_race}
count(clean_shot, v_race)
clean_shot$v_race <- recode_factor(clean_shot$v_race, 'AMERICAN INDIAN/ALASKAN NATIVE' = "Native",
   'ASIAN / PACIFIC ISLANDER' = "Asian", 
   'BLACK' = "Black", 
   'BLACK HISPANIC' = "Hispanic",
   'UNKNOWN' = "Unknown", 
   'WHITE' = "White", 
   'WHITE HISPANIC' = "Hispanic")

# Remove the handful of records that aren't in primary demo groups
clean_shot_5 <- subset(clean_shot, v_race != "Native")
clean_shot_4 <- subset(clean_shot_5, v_race != "Unknown")

# Now generate borough-specific data frames
Bronx_shot <- subset(clean_shot_4, bor == "BRONX")
Brooklyn_shot <- subset(clean_shot_4, bor == "BROOKLYN")
Manhattan_shot <- subset(clean_shot_4, bor == "MANHATTAN")
Queens_shot <- subset(clean_shot_4, bor == "QUEENS")
Staten_shot <- subset(clean_shot_4, bor == "STATEN ISLAND")

# Next, generate percentages for each borough
Bronx_pct <- count(Bronx_shot, v_race)
Bronx_pct$n <- Bronx_pct$n/sum(Bronx_pct$n)
Brooklyn_pct <- count(Brooklyn_shot, v_race)
Brooklyn_pct$n <- Brooklyn_pct$n/sum(Brooklyn_pct$n)
Manhattan_pct <- count(Manhattan_shot, v_race)
Manhattan_pct$n <- Manhattan_pct$n/sum(Manhattan_pct$n)
Queens_pct <- count(Queens_shot, v_race)
Queens_pct$n <- Queens_pct$n/sum(Queens_pct$n)
Staten_pct <- count(Staten_shot, v_race)
Staten_pct$n <- Staten_pct$n/sum(Staten_pct$n)
```
## Determining Borough Demographics

Now we gather some background demographic information by borough.

```{r get_demo_data}
demo_data <- read_csv("https://furmancenter.org/files/NDP/race_composition_data.csv")
clean_demo = select(demo_data, Geography, pop_hispanic_pct, pop_non_hispanic_asian_pct, pop_non_hispanic_black_pct, pop_non_hispanic_white_pct)
clean_demo <- clean_demo %>% rename(bor = Geography, hisp = pop_hispanic_pct, asian = pop_non_hispanic_asian_pct, black = pop_non_hispanic_black_pct, white = pop_non_hispanic_white_pct)

# translate % strings to numeric
clean_demo$hisp <- as.numeric(sub("%","",clean_demo$hisp))/100
clean_demo$asian <- as.numeric(sub("%","",clean_demo$asian))/100
clean_demo$black <- as.numeric(sub("%","",clean_demo$black))/100
clean_demo$white <- as.numeric(sub("%","",clean_demo$white))/100

# tidy these rows into borough-specific objects
NYCdf <- clean_demo[1,] %>% pivot_longer(cols = 2:5)
Bronxdf <- clean_demo[2,] %>% pivot_longer(cols = 2:5)
Brooklyndf <- clean_demo[3,] %>% pivot_longer(cols = 2:5)
Manhattandf <- clean_demo[4,] %>% pivot_longer(cols = 2:5)
Queensdf  <- clean_demo[5,] %>% pivot_longer(cols = 2:5)
StatenIslanddf <- clean_demo[6,] %>% pivot_longer(cols = 2:5)
```

## Comparing Shooting Victims to General Population

And now we can compare the racial profile of NYC shooting victims,
by borough, versus that borough's general demographics.

```{r plot_pies}
ShootLabels <- c('Asian', 'Black', 'Hispanic', 'White')
ShootColors <- c("blue", "brown", "green", "white")
FurmanLabels <- c('Hispanic', 'Asian', 'Black', 'White')
FurmanColors <- c("green", "blue", "brown", "white")
par(mfrow=c(1,2))
pie(Bronx_pct$n, ShootLabels, main="Bronx shooting victims", col=ShootColors)
pie(Bronxdf$value, FurmanLabels, main="Bronx demographics", col=FurmanColors)

par(mfrow=c(1,2))
pie(Brooklyn_pct$n, ShootLabels, main="Brooklyn shooting victims", col=ShootColors)
pie(Brooklyndf$value, FurmanLabels, main="Brooklyn demographics", col=FurmanColors)

par(mfrow=c(1,2))
pie(Manhattan_pct$n, ShootLabels, main="Manhattan shooting victims", col=ShootColors)
pie(Manhattandf$value, FurmanLabels, main="Manhattan demographics", col=FurmanColors)

par(mfrow=c(1,2))
pie(Queens_pct$n, ShootLabels, main="Queens shooting victims", col=ShootColors)
pie(Queensdf$value, FurmanLabels, main="Queens demographics", col=FurmanColors)

par(mfrow=c(1,2))
pie(Staten_pct$n, ShootLabels, main="Queens shooting victims", col=ShootColors)
pie(StatenIslanddf$value, FurmanLabels, main="Queens demographics", col=FurmanColors)

```
### Analysis

In every borough, gun violence victims were more likely to be Black than members of the general population of that borough. White and Asian members of the general population were less likely to be gunshot victims.

## Conclusion

The data indicates that, in every borough, Black residents are more likely to be gun victims than White and Asian residents. In Manhattan, a higher percentage of Hispanics were victims compared to the percentage of Hispanics who live there. Otherwise, that percentage was lower that the standard Hispanic demographic percentage.

Possible sources of bias include the systems used by New York City to discover and categorize shooting victims. It is also possible that some Asian and White victims decline to notify the authorities. My personal bias, based on media hype, was to expect a higher percentage of White victims, but that was not reflected in the data.
